An efficient composition of bidirectional programs by tupling and lazy updates

### 2019-09-27 paper/7.experiments.tex
split tex + compress
missing: abstract, intro, related work, conclusion, reference, appendix

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage{color}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{subfig}
\usepackage{lmodern}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}

\pgfplotsset{compat=1.11}

\setlength{\parindent}{0pt}
\renewcommand{\indent}{\hspace*{0pt}}

\newcommand{\tab}{\hspace*{5mm}}
\newcommand{\qtab}{\hspace*{5mm} \ \quad}

\newcommand{\sif}[3]{\text{if } #1 \text{ then } #2 \text{ else } #3}
\newcommand{\product}[2]{#1 \times #2}
\newcommand{\tuple}[2]{(#1 :: #2)}
\newcommand{\rearrs}[3]{RearrS \ #1 \ #2 \ #3}
\newcommand{\rearrv}[3]{RearrV \ #1 \ #2 \ #3}
\newcommand{\casebx}[4]{Case \ #1 \ #2 \ #3 \ #4}

\newcommand{\putbx}[3]{put \, [\![#1]\!] \ #2 \ #3}
\newcommand{\putbxinline}[1]{put \, [\![#1]\!]}
\newcommand{\getbx}[2]{get \, [\![#1]\!] \ #2}
\newcommand{\getbxinline}[1]{get \, [\![#1]\!]}

\newcommand{\putrev}[3]{put_{REV} \, [\![#1]\!] \ {#2} \ {#3}}
\newcommand{\getrev}[2]{get_{REV} \, [\![#1]\!] \ {#2}}

\newcommand{\pg}[3]{pg \, [\![#1]\!] (#2, #3)}
\newcommand{\pginline}[1]{pg \, [\![#1]\!]}

\newcommand{\cpg}[5]{cpg [\![#1]\!] (#2, #3, #4, #5)}
\newcommand{\cpginline}[1]{cpg \, [\![#1]\!]}

\newcommand{\kpg}[7]{kpg [\![#1]\!] (#2, #3, #4, #5, #6, #7)}
\newcommand{\kpginline}[1]{kpg \, [\![#1]\!]}

\newcommand{\xpg}[3]{xpg \, [\![#1]\!] (#2, #3)}
\newcommand{\xpginline}[1]{xpg \, [\![#1]\!]}

\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\newcolumntype{C}[1]{>{\centering\arraybackslash}p{#1}}
\newcolumntype{R}[1]{>{\raggedleft\arraybackslash}p{#1}}

\begin{document}
%
\title{An efficient composition of bidirectional programs
  by tupling and lazy updates 
%  tupling
%  accumulating updates = lazy updates
%  based on the idea of reversible computing  
  % \thanks{Supported by organization x.}
}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
%
\authorrunning{F. Author et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Princeton University, Princeton NJ 08544, USA \and
Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
\email{lncs@springer.com}\\
\url{http://www.springer.com/gp/computer-science/lncs} \and
ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
\email{\{abc,lncs\}@uni-heidelberg.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
The abstract should briefly summarize the contents of the paper in
15--250 words.

\keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}
%
%
%

% plan to pages
% abstraction & title & Introduction: 3pages
% miniBiGUL: 1.5page
% pg: 3pages
% cpg: 2page
% kpg: 1page
% xpg:0.5page
% evaluation:2pages
% related work:1page
% conclusion:1page (including reference)
% page limit : 15

Schedule
\begin{itemize}

\item Introduction (story), Contribution, positions, examples, beginning of September?
\item zero draft : mid of September

\item deadline of PEPM2020 (October 18th)
\item ESOP (October 24th)
\item FLOPS (the middle of November??)
\item Need to change the style file of latex..
\item []
\end{itemize}

Points to discuss
\begin{itemize}
\item Introduction (story)
\item Contributions
\item Related work?
\item One week to revise current version.
  After that, Prof. Hu and Robert, could you send comments?

%\item TODO: %relationship between result (original vs \{cpg, kpg, xpg\})
%\item right assoc \& Recursive example?
%\item current foldr is not efficient? Are there any efficient foldr? Question to Prof. Hu.
%\item Language to show semantics: fi?, $f^{-1}$, $cond_{sv}$ and $cond_{s}$
%\item A good simple example of bx: Used in explanation of BiGUL and evaluation of pg.
%\item Related work: suggest?
%\item another pg ?? should be compared?
\end{itemize}

\section{Introduction}

%\begin{itemize}
%\item Importance of BX, BX is a solution of view update problem in database.
%\item goodness of 
%\item Explanation of put-based BX: BiGUL.
%\item Current status of BiGUL: Fastest BX language in the world
%\item But there is a problem: Efficiency of compose evaluation. The current implementation of BiGUL does not save the intermediate states, the number of get is quadratic. This is not good.
%\item To solve this problem we use an idea: introduce pg : combination of put and get. Then, no information will be lost in a specific condition.
%  an idea from reversible computation: not to lose any information.
%\item We extend pg with several ideas to produce faster implementation.
%\item 
%\end{itemize}

%\begin{itemize}
%\item Importance of BX, BX is a solution of view update problem in database.
%\item goodness of 
%\item Explanation of put-based BX: BiGUL.
%\item Current status of BiGUL: Fastest BX language in the world
%\item But there is a problem: Efficiency of compose evaluation. The current implementation of BiGUL does not save the intermediate states, the number of get is quadratic. This is not good.
%\item To solve this problem we use an idea: introduce pg : combination of put and get. Then, no information will be lost in a specific condition.
%  an idea from reversible computation: not to lose any information.
%\item We extend pg with several ideas to produce faster implementation.
%\item 
%\end{itemize}

In software, there are strong demands for synchronizing the data. In database community it is known as ``view update problem'' and researched for a long time \cite{}.
% A Survey to View Update Problem 
As a solution for this problem, bidirectional transformation is introduced \cite{}. It enables the evaluation of source updating: ``put'' produces the updated source from the original source and the updated view.
% The first paper of bidirectional transformation

There are two types of bidirectional transformation. One is get-based bidirectional transformation \cite{} that receives the definition of $get$ and provides $put$. A problem of this approach is that it is possible to derive many correct $put$s. To choose one put, hulistics and programmer's intentions are needed. Another approach is put-based bidirectional transformation \cite{} that received the definition of $put$ and provide $get$. From a $put$ we can derive an unique $get$.

BiGUL \cite{} is one of put-based bidirectional transformation language. It is fastest bidirectional language in the world now. However, it contains a problem when the BiGUL programs including many compositions. BiGUL does not store the intermediate states when $get$s are re-evaluated so many times. In this case the number of get is quadratic.

\vspace{5mm}

Figure that shows many gets re-evaluation

\vspace{5mm}

% general, concrete problem
% get-based also have the same problem. even in lenses. more general problem in
% The example should be from lenses.
% traversal, not exactly the same computation

To solve this problem and make BiGUL evaluation efficient, we use two techniques. The first technique is tupling, well-known technique for efficiency \cite{}. Its basic idea is to reduce the common computations by combining two evaluations. In our work, we combines the evaluation of $put$ and $get$, and introduce $pg$. Thanks to tupling, we do not lose information like intermediate states.
The second technique is lazy update and lazy computation. $pg$ itself includes some inefficiency. For removing the inefficiency we do the update and computation lazily.

The rest of this paper is the following. We introduce a subset of BiGUL, miniBiGUL, and show its syntax, semantics, and properties in Section 2. Combining the semantics of put and get in Section 2, we construct $pg$ and show that it is self-inverse in Section 3.
We apply three optimizations to $pg$ in sequence. The first one is keeping update of BX lazily, and we obtain $cpg$ in Section 4. The second one is lazy evaluation to avoid redundant computations, and we obtain $kpg$ in Section 5. The third one is combination of $pg$ and $kpg$ in Section 6, because they are strong for different type of programs. In Section 7, we show experiemntal results of $pg$, $cpg$, $kpg$, $xpg$ and miniBiGUL.
In Section 8 we discuss about related work, and conclude in Section 9.

% Story ->
% reversible is needed not to lose any information -> pg is introduced.
% pg is not good for lassoc -> cpg (xpg_2) is introduced.
% cpg can be improved -> kpg(xpg) is introduced.
% kpg is slower than pg for rassoc. combined version -> xpg.

% Evaluation result: Comparison with OCaml implementation of minBiGUL

%\begin{itemize}
%\item Importance of BX, BX is a solution of view update problem in database.
%\item Explanation of put-based BX: BiGUL.
%\item Current status of BiGUL: Fastest BX language in the world
%\item But there is a problem: Efficiency of compose evaluation. The current implementation of BiGUL does not save the intermediate states, the number of get is quadratic. This is not good.
%\item To solve this problem we use an idea: introduce pg : combination of put and get. Then, no information will be lost in a specific condition.
%  an idea from reversible computation: not to lose any information.
%\item We extend pg with several ideas to produce faster implementation.
%\item 
%\end{itemize}

% get & put independently, but for efficient computation
% we need to tuple them to help composition
% 2 or 3 examples to show tupling solves the problem.
% other examples are for showing kpg and cpg is needed.

\subsection{Contribution}

\begin{itemize}
\item Improvement of evaluation efficiency
  \begin{itemize}
  \item Quadratic to exponential? 
  \item The introduced extentions (maybe) use more memory than the original approach, but they are faster than the original.
  \end{itemize}
\item Get and put more tight

\item (Tupling introduces some constructions)
  \begin{itemize}
  \item A pair of put and get is self-inverse (involution).
  \end{itemize}
\item careful lazyness is needed in our implementation for efficiency
%  lazy update \& lazy computation are also good optimization for BX.
%\item Reveal a part of relationship between BX and Reversible.
%  \begin{itemize}
%  \item Very well-behaved (no adaptive-case) \& The same evaluation path $\to$ No garbage Reversible Computation.
%  \item 
%  \item If we can obtain good parameter (dummy) for a BiGUL program,
%    put(s, dummy) can be a complement function.
  \end{itemize}

%\subsection{Construction of this paper}

%\begin{itemize}
%\item BiGUL 
%\item pg (pair of put and get)
%\item cpg (extention of pg. Keep eval information for reusing)
%\item kpg (extention of cpg. Lazy evaluation)
%\item xpg (combination of pg and kpg)
%\item Experiment
%\item Related work
%\item Conclusion and Future work
%\end{itemize}

\section{Bidirectional Programming Language: minBiGUL}

MinBiGUL, our target language in this paper, is a subset of BiGUL [?] which is a simple yet powerful putback-based bidirectional language. BiGUL supports two transformations: a forward transformation $get$ producing a view from a source and a backward transformation $put$ taking a source and a modified view to produce an updated source. Intuitively, if we have a BiGUL program $bx$, these two transformations are following functions:\\
    $\tab \getbxinline{bx} = fun \ s \to v \\
    \tab \putbxinline{bx} = fun \ s \ v \to v$

BiGUL is well-behaved since two functions $\putbxinline{bx}$ and $\getbxinline{bx}$ satify the round-trip laws as follows:\\
    $\tab \putbx{bx}{s}{(\getbx{bx}{s})} = s \qquad [\text{GETPUT}]\\
    \tab \getbx{bx}{(\putbx{bx}{s}{v})} = v \qquad [\text{PUTGET}]$

The GETPUT law means that if there is no change to the view, there should be no change to the source. The PUTGET law means that we can recover the modified view by applying the forward transformation to the updated source.

MinBiGUL inherited from BiGUL also supports transformations $put$ and $get$ which are satify two above laws. In addition, we restrict adaptive cases of BiGUL on minBiGUL. Then $put$ and $get$ satify one more law, PUTPUT, like the following:\\
    $\tab \putbx{bx}{(\putbx{bx}{s}{v'})}{v} = \putbx{bx}{s}{v} \qquad [\text{PUTPUT}]$

The PUTPUT law means that a source update should overwrite the effect of previous source updates.

Due to the satisfaction of three laws, GETPUT, PUTGET and PUTPUT, minBiGUL is very well-behaved.

In the remaining of this section, we will present about the syntax and semantics of minBiGUL through formal definitions and examples.

\subsection{Syntax}

The syntax of minBiGUL is briefly written as follows:

$bx := Skip \ h \\ 
    \qtab | \ Replace \\ 
    \qtab | \ Prod \ bx_1 \ bx_2 \\ 
    \qtab | \ RearrS \ f_1 \ f_2 \ bx \\
    \qtab | \ RearrV \ g_1 \ g_2 \ bx \\
    \qtab | \ Case \ cond_{sv} \ cond_{s} \ bx_1 \ bx_2 \\ 
    \qtab | \ Compose \ bx_1 \ bx_2$

A minBiGUL program may be either a skip of a function or a replacement or a product of two minBiGUL programs or a source / view rearrangement or a case combinatator without adaptive cases or a composition of some minBiGUL programs.

For source / view rearrangement, BiGUL use just one lambda expression to express how to deconstruct as well as reconstruct data. It is a kind of bijection. However, to be able to implement it in OCaml, the environment used for developing minBiGUL and solutions in the paper, we need give two functions which one is the inverse of the other. In the above syntax, $f_2 = f_1^{-1}$ and $g_2 = g_1^{-1}$.

To help make demonstration more direct, we provide the following alternatives representation:

    $\tab Prod \ bx_1 \ bx_2 \equiv bx_1 \times bx_2$\\
    $\tab Compose \ bx_1 \ bx_2 \equiv bx_1 \circ bx_2$

In general, $\circ$ has a higher priority than $\times$. Since only consider a product of two minBiGUL programs is considered, it is unnecessary to define the associativity precedence of $\times$. The one of $\circ$ may be left or right, however we do not declare the default.

In the next part, we will introduce the semantics of minBiGUL.

\subsection{Semantics}

\begin{definition}
    $\putbx{bx}{s}{v}$

    $\putbx{Skip \ h}{s}{v} = \sif{h \ s = v}{s}{\text{fail}}$

    $\putbx{Replace}{s}{v} = v$

    $\putbx{\product{bx_1}{bx_2}}{(s_1, s_2)}{(v_1, v_2)} =\\
        \tab ((\putbx{bx_1}{s_1}{v_1}),(\putbx{bx_2}{s_2}{v_2}))$

    $\putbx{\rearrs{f_1}{f_2}{bx}}{s}{v} = f_2 \ (\putbx{bx}{(f_1 \ s)}{v})$

    $\putbx{\rearrv{g_1}{g_2}{bx}}{s}{v} = \putbx{bx}{s}{(g_1 \ v)}$

    $\putbx{\casebx{cond_{sv}}{cond_{s}}{bx_1}{bx_2}}{s}{v} = \\
        \tab \text{if} \ {cond_{sv} \ s \ v} \\
        \tab \text{then} \ s' \Leftarrow \putbx{bx_1}{s}{v} \\
        \tab \text{else} \ s' \Leftarrow \putbx{bx_2}{s}{v} \\
        \tab \text{fi} \ cond_{s} \ s'; \ \text{return} \ s'$

    $\putbx{bx_1 \circ bx_2}{s}{v} = \putbx{bx_1}{s}{(\putbx{bx_2}{(\getbx{bx_1}{s})}{v})}$
\end{definition}

\begin{definition}
    $\getbx{bx}{s}$

    $\getbx{Skip \ h}{s} = h \ s$

    $\getbx{Replace}{s} = s$

    $\getbx{\product{bx_1}{bx_2}}{(s_1,s_2)} = \\
        \tab ((\getbx{bx_1}{s_1}),(\getbx{bx_2}{s_2}))$

    $\getbx{\rearrs{f_1}{f_2}{bx}}{s} = \getbx{bx}{(f_1 \ s)}$

    $\getbx{\rearrv{g_1}{g_2}{bx}}{s} = g_2 \ (\getbx{bx}{s})$

    $\getbx{\casebx{cond_{sv}}{cond_{s}}{bx_1}{bx_2}}{s} =\\
        \tab \text{if} \ {cond_{s} \ s}\\
        \tab \text{then} \ v' \Leftarrow \getbx{bx_1}{s}\\
        \tab \text{else} \ v' \Leftarrow \getbx{bx_2}{s}\\
        \tab \text{fi} \ {cond_{sv} \ s \ v'}; \ \text{return} \ v'$

    $\getbx{bx_1 \circ bx_2}{s} = \getbx{bx_2}{(\getbx{bx_1}{s})}$
\end{definition}

In defintions 1 and 2, we use if-then-else-fi statements to express semantics of $\putbxinline{Case}$ and $\getbxinline{Case}$. Statement (if $E_1$ then $X_1$ else $X_2$ fi $E_2$) means if the test $E_1$ is true, the statement $X_1$ is executed and the assertion $E_2$ must be true, otherwise, i.e. $E_2$ is false, the statement $X_2$ is executed and the assertion $E_2$ must be false. If the values of $E_1$ and $E_2$ are distinct, the if-then-else-fi structure is undefined. We can write the equivalent if-then-else statement as follows:

$\tab \text{if } E_1 \text{ then } X_1 \text{ else } X_2 \text{ fi } E_2; S\\
\tab \equiv \text{if }E_1 = true \text{ then}\\ 
    \qtab \tab X_1;\\ 
    \qtab \tab \text{if } E_2 = true \text{ then } S \text{ else assert } false;\\
    \qtab \text{else}\\ 
    \qtab \tab X_2;\\ 
    \qtab \tab \text{if } E_2 = false \text{ then } S \text{ else assert } false;$

The if-then-else-fi statement is useful to describe many functions related to $Case$ in this paper.

Next, let's take a look at some examples to better understand about minBiGUL. We start with quite obivious things as follows:\\
$\putbx{Skip \ (\lambda x.(x*x))}{10}{100} = 10$\\
$\putbx{Skip \ (\lambda \_.())}{1}{()} = 1$\\ 
$\getbx{Skip \ (\lambda x.(x*x))}{10} = 100$\\
$\getbx{Skip \ (\lambda \_.())}{1} = ()$\\
$\putbx{Replace}{1}{100} = 100$\\
$\getbx{Replace}{1} = 1$\\

Now, let's consider the definition of $phead$ in BiGUL [?]:

$\tab phead :: Show \ s \Rightarrow BiGUL \ [s] \ s\\
\tab phead = \$(RearrS \ [\![\lambda(s:ss) \to (s,ss)]\!])\$\\
    \qtab \qtab \quad \$(RearrV \ [\![\lambda v \to (v,())]\!])\$\\
        \qtab \qtab \qquad Replace \ `prod\text{'} \ (Skip \ (const \ ()))$

The above program rearranges the source, a non-empty list, to a pair of its head element $s$ and its tail $ss$, and the view to a pair $(v, ())$, then we can use $v$ to replace $s$ and $()$ to keep $ss$. $\putbx{phead}{s_0}{v_0}$ returns a list whose head is $v_0$ and tail is the tail of $s_0$. $\getbx{phead}{s_0}$ returns the head of $s_0$. For instance, $\putbx{phead}{[1,2,3]}{100} = [100,2,3]$ and $\getbx{phead}{[1,2,3]} = 1$.\\

In minBiGUL, we can express $phead$ as follows:

$\tab phead = RearrS \ f_1 \ f_2 \ bx_s \ \text{where:}\\
    \qtab f_1 = \lambda (s::ss).(s,ss)\\
    \qtab f_2 = \lambda (s,ss).(s::ss)\\
    \qtab bx_s = RearrV \ g_1 \ g_2 \ bx_v \ \text{where:}\\
        \qtab \qtab g_1 = \lambda v.(v,())\\
        \qtab \qtab g_2 = \lambda (v,()).v\\
        \qtab \qtab bx_v = \product{Replace}{(Skip \ (\lambda \_.())}$\\

Then, we can evaluate $\putbxinline{phead}$ and $\getbxinline{phead}$ based on their definitions:

$\putbx{phead}{[1,2,3]}{100}\\
\tab = \putbx{\rearrs{f_1}{f_2}{bx_s}}{[1,2,3]}{100}\\
\tab = f_2 \ (\putbx{bx_s}{(f_1 \ [1,2,3])}{100})\\
\tab = f_2 \ (\putbx{\rearrv{g_1}{g_2}{bx_v}}{(1,[2,3])}{100})\\
\tab = f_2 \ (\putbx{bx_v}{(1,[2,3])}{(g_1 \ 100}))\\
\tab = f_2 \ (\putbx{\product{Replace}{(Skip \ (\lambda \_.()))}}{(1,[2,3])}{(100,())})\\
\tab = f_2 \ ((\putbx{Replace}{1}{100}), (\putbx{(Skip \ (\lambda \_.()))}{[2,3]}{()}))\\
\tab = f_2 \ (100, [2,3])\\
\tab = [100,2,3]$\\

$\getbx{phead}{[1,2,3]}\\
\tab = \getbx{\rearrs{f_1}{f_2}{bx_s}}{[1,2,3]}\\
\tab = \getbx{bx_s}{(f_1 \ [1,2,3])}\\
\tab = \getbx{\rearrv{g_1}{g_2}{bx_v}}{(1,[2,3])}\\
\tab = g_2 \ (\getbx{bx_v}{(1,[2,3])})\\
\tab = g_2 \ (\getbx{\product{Replace}{(Skip \ (\lambda \_.()))}}{(1,[2,3])})\\
\tab = g_2 \ ((\getbx{Replace}{1}), (\getbx{(Skip \ (\lambda \_.()))}{[2,3]}))\\
\tab = g_2 \ (1,())\\
\tab = 1$\\

If we wanna update the head element of the head element of a list of lists by using the view, we can define a composition like $phead \circ phead$. For example:

    \tab $\putbx{phead \circ phead}{[[1,2,3],[\,],[4,5]]}{100} = [[100,2,3],[\,],[4,5]]$\\
    \tab $\getbx{phead \circ phead}{[[1,2,3],[\,],[4,5]]} = 1$\\

We distingwish two $phead$s in the composition by indices. According to definitions of $\putbxinline{bx}$ and $\getbxinline{bx}$, we have:

$\getbx{phead_1 \circ phead_2}{[[1,2,3],[\,],[4,5]]}\\
    \tab = \getbx{phead_2}{(\getbx{phead_1}{[[1,2,3],[\,],[4,5]]})}\\
    \tab = \getbx{phead_2}{[1,2,3]}\\
    \tab = 1$\\

$\putbx{phead_1 \circ phead_2}{[[1,2,3],[\,],[4,5]]}{100}\\
    \tab = \putbx{phead_1}{[[1,2,3],[\,],[4,5]]}{(\putbx{phead_2}{(\getbx{phead_1}{[[1,2,3],[\,],[4,5]]})}{100})}\\
    \tab = \putbx{phead_1}{[[1,2,3],[\,],[4,5]]}{(\putbx{phead_2}{[1,2,3]}{100})}\\
    \tab = \putbx{phead_1}{[[1,2,3],[\,],[4,5]]}{[100,2,3]}\\
    \tab = [[100,2,3],[\,],[4,5]]$\\

One more example we would like to present is $replaceAll$ whose definition in BiGUL as follows:\\
$\tab replaceAll :: (Eq \ s, Show \ s) \Rightarrow BiGUL \ [s] \ s \\ 
\tab replaceAll = Case \ [\\
    \qtab \$(normal \ [\![\lambda s \ v \to length \ s == 1]\!][\![\lambda s \to length \ s == 1]\!]) \\ 
    \qtab \Longrightarrow \$(rearrS \ [\![\lambda [x] \to x]\!]) Replace,\\
    \qtab \$(normal \ [\![\lambda s \ v \to length \ s > 1]\!][\![\lambda s \to length \ s > 1]\!]) \\ 
    \qtab \Longrightarrow \$(rearrS \ [\![\lambda (x:xs) \to (x,xs)]\!])\$\\ 
    \qtab \qtab \$(rearrV \ [\![\lambda v \to (v,v)]\!])\$\\ 
    \qtab \qtab Replace \ `prod\text{'} \ replaceAll\\
\tab ]$\\

The following expresses $replaceAll$ in minBiGUL:\\
$\tab replaceAll = Case \ cond_{sv} \ cond_{s} \ bx_1 \ bx_2 \ \text{where:}\\ 
    \qtab cond_{sv} = \lambda s.\lambda v.length \ s == 1\\
    \qtab cond_{s} = \lambda s.length \ s == 1\\ 
    \qtab bx_1 = RearrS \ (\lambda [x] \to x) \ (\lambda x \to [x]) \ Replace\\ 
    \qtab bx_2 = RearrS \ (\lambda (x::xs) \to (x,xs)) \ (\lambda (x,xs) \to (x::xs)) \ bx_s \ \text{where:}\\ 
        \qtab \tab bx_s = RearrV \ (\lambda v \to (v,v)) \ (\lambda (v,v) \to v) \ bx_v \ \text{where:}\\
            \qtab \qtab bx_v = \product{Replace}{replaceAll}$

We can also evaluate $\putbxinline{replaceAll}$ and $\getbxinline{replaceAll}$ similarly to the previous examples if we pay attention to conditions $cond_sv$ and $cond_s$ as well as the if-then-esle-fi structure. $\putbxinline{replaceAll}$ uses the view to update all elements in the source list, while $\getbxinline{replaceAll}$ returns the an element of the source list with same elements. For example, $\putbx{replaceAll}{[1,2,3]}{100} = [100,100,100]$ and $\getbx{replaceAll}{[1,1,1]} = 1$.

\section{pg}
\subsection{Self-inverse function: pg}

In minBiGUL or even BiGUL, when evaluating a composition of many programs, $get$s are re-evaluated so many times since no intermediate state is stored during the evaluation. This is a kind of information loss. One question is whether it is possible to calculate such programs without losing information. And the answer is yes. It comes from the the idea of reversible computation where all information during the evaluation need to be kept. In minBiGUL, we can do that by tupling $put$ and $get$. A pair of a source and a view is accepted as the input of a function named $pg$ to produce a new pair that contains the actual result of a corresponding minBiGUL program.

\begin{definition}
    $\pg{bx}{s}{v} = (\putbx{bx}{s}{v}, \getbx{bx}{s})$
\end{definition}

$pg$ is an involution that is self-inverse. An involution is a function $f$ that satifies $f(f(x)) = x \text{ for all } x \text{ in the domain of } f$.

\begin{proof}
$pg [\![bx]\!] \ (\pg{bx}{s}{v}) \\
    \tab = \pg{bx}{(\putbx{bx}{s}{v})}{(\getbx{bx}{s})} \quad [pg \text{ definition}] \\
    \tab = (put \ [\![bx]\!] \ (\putbx{bx}{s}{v}) \ (\getbx{bx}{s}), \getbx{bx}{(\putbx{bx}{s}{v})})  \quad [pg \text{ definition}] \\
    \tab = (put \ [\![bx]\!] \ (\putbx{bx}{s}{v}, \getbx{bx}{s}), v) \quad [\text{PUTGET}] \\
    \tab = (put \ [\![bx]\!] \ (s, \getbx{bx}{s}), v) \quad [\text{PUTPUT}] \\
    \tab = (s, v) \quad [\text{GETPUT}]$
\end{proof}

\subsection{Construction of pg}

$\pg{Skip \ h}{s}{v} \\
    \tab = (\sif{h \ s = v}{s}{\text{fail}}, h \ s) \\
    \tab = \sif{h \ s = v}{(s, h \ s)}{\text{fail}} \\
    \tab = \sif{h \ s = v}{(s, v)}{\text{fail}}$

There is a trick in the construction of $\pginline{Skip \ h}$. The first equality is simply based on the definitions of $pg$, $\putbxinline{Skip \ h}$ and $\getbxinline{Skip \ h}$. The second one tuples two results of $put$ and $get$ in the body of the if-expression. And the last one is quite obivious. What we call the trick here is in the second equality where in some cases, the result of $pg$ may be fail although there is no fail when evaluating $\getbxinline{Skip \ h}$.\\

$\pg{Replace}{s}{v}\\
    \tab = (v, s)$

$\pg{\product{bx_1}{bx_2}}{(s_1,s_2)}{(v_1,v_2)} \\
    \tab = ((\putbx{bx_1}{s_1}{v_1}),(\putbx{bx_2}{s_2}{v_2}), (\getbx{bx_1}{s_1}), (\getbx{bx_2}{s_2})) \\
    \tab = (s_1, v_1) \Leftarrow \pg{bx_1}{s_1}{v_1}; \\
        \qtab (s_2, v_2) \Leftarrow \pg{bx_2}{s_2}{v_2}; \\
        \qtab \tab ((s_1,s_2), (v_1,v_2))$

$\pg{\rearrs{f_1}{f_2}{bx}}{s}{v} \\
    \tab = (f_2 \ (\putbx{bx}{(f_1 \ s)}{v}), \getbx{bx}{(f_1 \ s)}) \\
    \tab = (s, v) \Leftarrow \pg{bx}{f_1 \ s}{v};\\
        \qtab \tab (f_2 \ s, v)$

$\pg{\rearrv{g_1}{g_2}{bx}}{s}{v} \\
    \tab = (\putbx{bx}{s}{(g_1 \ v)}, g_2 \ (\getbx{bx}{s})) \\
    \tab = (s, v) \Leftarrow \pg{bx}{s}{g_1 \ v}; \\
        \qtab \tab (s, g_2 \ v)$\\

Constructions of $pg$ for the replacement, the product and the source / view rearrangements are very clear when just paring $put$ and $get$ respectively, then doing basic changes.\\

$\pg{\casebx{cond_{sv}}{cond_{s}}{bx_1}{bx_2}}{s}{v} \\
    \tab = (\text{if} \ {cond_{sv} \ s \ v} \\
    \qtab \text{then} \ s' \Leftarrow \putbx{bx_1}{s}{v} \\
    \qtab \text{else} \ s' \Leftarrow \putbx{bx_2}{s}{v} \\
    \qtab \text{fi} \ cond_{s} \ s'; \text{ return } s', \\
    \qtab \quad \text{if} \ {cond_{s} \ s}\\
    \qtab \quad \text{then} \ v' \Leftarrow \getbx{bx_1}{s}\\
    \qtab \quad \text{else} \ v' \Leftarrow \getbx{bx_2}{s}\\
    \qtab \quad \text{fi} \ {cond_{sv} \ s \ v'}; \text{ return } v') \\
    \tab = \text{if} \ {cond_{sv} \ s \ v} \ \&\& \ {cond_{s} \ s} \\
        \qtab \text{then} \ (s', v') \Leftarrow \pg{bx_1}{s}{v}\\
        \qtab \text{else} \ (s', v') \Leftarrow \pg{bx_2}{s}{v}\\
        \qtab \text{fi} \ cond_{s} \ s' \ \&\& \ cond_{sv} \ s \ v'; \text{ return } (s',v')$\\

>>>>

\section{Experiments}

This section describes the experiments involving $minBiGUL$, $pg$, $cpg$, $kpg$ and $xpg$. The evaluation time and the memory allocation are considered in each test.

\subsection{Experiment environment \& Test cases}
We implemented 5 approaches with OCaml 4.07.1 in the same environment as follows: macOS 10.14.6, processor Intel Core i7 (2.6 GHz), RAM 16 GB 2400 MHz DDR4. The OCaml runtime system options and garbage collection parameters are set as default.

We also conducted tests on 5 cases, including many composition styles such as left or right associative, non-recursive or recursive. Table \ref{tab:test-cases} shows more details on these cases.

\begin{table}[hbt!]
    \centering
    \caption{Test cases}
    \label{tab:test-cases}
    \begin{tabular*}{\textwidth}{|c @{\extracolsep{\fill}}|l|l|l|c|c|c|c|}
        \hline
        \multirow{2}{*}{No} & \multicolumn{1}{c|}{\multirow{2}{*}{Name}} & \multicolumn{1}{c|}{\multirow{2}{*}{Associative}} & \multicolumn{1}{c|}{\multirow{2}{*}{Recursive}} & \multicolumn{2}{c|}{Input} & \multicolumn{2}{c|}{Output} \\ \cline{5-8} 
        & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{$s_0$} & \multicolumn{1}{c|}{$v_0$} & \multicolumn{1}{c|}{$s_r$} & \multicolumn{1}{c|}{$v_r$} \\ \hline
        1 & lassoc-comp-replace & left & no & 1 & 100 & 100 & 1 \\ \hline
        2 & rassoc-comp-replace & right & no & 1 & 100 & 100 & 1 \\ \hline
        3 & lassoc-comp-phead & left & no & $\underbrace{[[\ldots[1]\ldots]]}_{\text{n+1 times}}$ & 100 & $\underbrace{[[\ldots[100]\ldots]]}_{\text{n+1 times}}$ & 1 \\ \hline
        4 & rassoc-comp-phead & right & no & $\underbrace{[[\ldots[1]\ldots]]}_{\text{n+1 times}}$ & 100 & $\underbrace{[[\ldots[100]\ldots]]}_{\text{n+1 times}}$ & 1 \\ \hline
        5 & breverse & left/right & yes & [1..n] & [1..n] & [n..1] & [n..1] \\ \hline
    \end{tabular*}
\end{table}

In table \ref{tab:test-cases}, $s_r$ and $v_r$ are respectively updated source and view which are produced by applying $put$ and $get$ on original source $s_0$ and original view $v_0$. This means: $s_r = \putbx{bx}{s_0}{v_0}$ and $v_r = \getbx{bx}{s_0}$, where $bx$ is one of the 5 cases mentioned in the table. Note that the results $s_r$ and $v_r$ do not depend on the associative style of the composition. The first four test cases simply use $n$ compose operators to make a composition of $n + 1$ similar $BX$ programs which are non-recursive. For example, \textit{lassoc-comp-replace}, left-associative composition of $Replace$s, looks like $(\ldots((Replace \circ Replace) \circ Replace) \circ \ldots \circ Replace) \circ Replace$, while \textit{rassoc-comp-replace}, right-associative composition of $Replace$s, is like $Replace \circ (Replace \circ (Replace \circ \ldots \circ (Replace \circ Replace)\ldots))$, where there are $n + 1 \ Replace$s in each case. The last test case, \textit{breverse}, is defined in terms of \textit{bFoldr}, a putback function for \textit{foldr} which is an important higher-order function on lists. This case is more natural than the other since \text{bFoldr} is a Case analysis whose a branch contains a composition $bx_1 \circ bx_2$ where $bx_1$ is a product of a simple function and $bFoldr$. It will look like a composition of more BX programs at some point if we slow down the evaluation. Note that $\circ$ has a higher priority than $\times$. This leads to seemingly impossible to transform \textit{bReverse} as well as \textit{bFoldr} from left-associative style to right-associative style.
